---
title: "Ethiopia 2024-25 Contraceptive Calendar Analysis"
subtitle: "Reproducible initiation, discontinuation, reason, and reinitiation workflow"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
    code_folding: show
    theme: flatly
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)

rmd_file <- tryCatch(
  normalizePath(knitr::current_input(dir = TRUE), mustWork = TRUE),
  error = function(e) NA_character_
)
project_dir <- if (!is.na(rmd_file)) {
  dirname(dirname(rmd_file))
} else {
  normalizePath(file.path(".."))
}
options(ethiopia_discontinuation_project_dir = project_dir)

input_file <- file.path(project_dir, "Data", "Ethiopia", "ETIR8AFL.dta")
derived_dir <- file.path(project_dir, "Analyses", "derived")
output_dir <- file.path(project_dir, "output")
code_dir <- dirname(rmd_file)
```

# Purpose and outputs

This R Markdown file is the reproducible entry point for the Ethiopia
contraceptive-calendar analysis. It documents and executes the complete
workflow used to generate the analytical CSV files and the self-contained HTML
report.

The final dashboard is written to:

```text
output/ethiopia_discontinuation_3_6_9_12.html
```

The analysis covers:

- observed contraceptive method starts, initiation mix, and initiation rate;
- all-cause discontinuation at 3, 6, 9, and 12 months;
- competing-risk cumulative incidence by discontinuation reason;
- the distribution of reasons among discontinued episodes; and
- time to contraceptive reinitiation following a non-pregnancy-related
  discontinuation.

# Input data

The required input is the Ethiopia 2024-25 DHS Individual Recode Stata file:

```{r check-input}
if (!file.exists(input_file)) {
  stop(
    paste0(
      "Missing Data/Ethiopia/ETIR8AFL.dta. Place the Ethiopia IR file in ",
      "Data/Ethiopia/ and rerun."
    )
  )
}

file.info(input_file)[, c("size", "mtime")]
```

Only the variables needed for calendar reconstruction are initially read:

```{r read-data-example, eval=FALSE}
library(haven)

calendar <- read_dta(
  input_file,
  col_select = c(
    caseid, # anonymized respondent identifier
    v005,   # DHS woman sampling weight
    v008,   # interview century-month code
    v017,   # first calendar month in century-month-code form
    v018,   # position of the calendar at interview
    v019,   # calendar length
    vcal_1, # monthly pregnancy, birth, non-use, and method states
    vcal_2  # reason recorded when a method-use episode ends
  )
)

calendar$weight <- as.numeric(calendar$v005) / 1e6
```

The executable implementation is maintained in
`Code/replicate_table_7_11.R`. Keeping one calculation engine prevents the code
shown in the documentation from drifting away from the code used for results.

# Converting the calendar into episodes

## Monthly calendar orientation

The contraceptive calendar is stored as one character per observed month.
The code first pads the calendar strings consistently, identifies the observed
positions for each respondent, and converts each position to a century-month
code relative to `v008`.

```{r calendar-orientation-example, eval=FALSE}
start_position <- as.integer(v018)
end_position <- min(80L, start_position + as.integer(v019) - 1L)
positions <- seq.int(end_position, start_position, by = -1L)

method_state <- substring(vcal_1, positions, positions)
reason_state <- substring(vcal_2, positions, positions)
month_cmc <- as.integer(v008) - (positions - start_position)
```

## Episode construction

Consecutive months with the same `vcal_1` code are combined into continuous
events. For each run, the analysis retains:

- respondent identifier;
- episode start and end century-month codes;
- episode duration;
- method code;
- discontinuation-reason code at the episode endpoint; and
- DHS sampling weight.

```{r episode-construction-example, eval=FALSE}
runs <- rle(method_state)
run_end <- cumsum(runs$lengths)
run_start <- c(1L, head(run_end, -1L) + 1L)

# For each run j:
idx <- run_start[j]:run_end[j]
episode_start <- min(month_cmc[idx])
episode_end <- max(month_cmc[idx])
episode_duration <- length(idx)

# The reason is stored in the most recent month of the episode.
reason_code <- reason_state[max(idx)]
```

Birth, pregnancy, termination, and non-use runs remain available for temporal
sequencing but are excluded from the contraceptive discontinuation risk set.

# DHS discontinuation analysis window

The standard DHS rules are applied to the period 3-62 months before interview:

1. Drop events already underway when the oldest calendar month begins because
   their true start and duration are unknown.
2. Keep contraceptive method-use episodes and exclude non-use, pregnancy,
   birth, and termination runs from the discontinuation risk set.
3. Treat discontinuations during the final three months as censored.
4. Exclude episodes starting during those final three months.
5. Exclude episodes that both began and ended before the 62-month boundary.
6. Allow an episode beginning before the 62-month boundary to contribute when
   it remained in use into the analysis window, using delayed entry.

```{r dhs-window-example, eval=FALSE}
months_start_to_interview <- interview_cmc - event_start
months_end_to_interview <- interview_cmc - event_end

discontinued <- as.integer(reason_code != 0)
discontinued[months_end_to_interview < 3] <- 0L

entry <- ifelse(
  months_start_to_interview >= 63,
  months_start_to_interview - 62,
  0
)

keep <- months_start_to_interview >= 3 &
  !(months_start_to_interview > 62 & months_end_to_interview > 62)
```

The denominator for a discontinuation probability is therefore the weighted
life-table risk set of eligible use episodes, not only the episodes that
discontinued. Episodes leave the risk set when they discontinue or are
censored.

# Method and reason groupings

The report uses the DHS aggregate method categories:

- Injectables
- Implants
- Pill
- Emergency contraception
- Other
- All methods

Reason codes are grouped into nine mutually exclusive causes:

1. Method failure  
2. Desire to become pregnant  
3. Other fertility-related reasons  
4. Changes in menstrual bleeding  
5. Other side effects or health concerns  
6. Wanted a more effective method  
7. Other method-related reasons  
8. Husband or partner disapproved  
9. Other reasons

The precise survey-specific mappings are defined and documented in
`Code/replicate_table_7_11.R`.

# Initiation measures

An observed start is the beginning of a method-use episode for which the
preceding calendar state is observed. Episodes already underway at the oldest
calendar boundary are not counted as starts.

Two complementary summaries are produced:

```{r initiation-definitions, echo=FALSE, results='asis'}
cat(
  "- **Initiation mix:** weighted starts of a method divided by weighted starts",
  "of all methods. This describes the composition of observed starts.\n",
  "- **Initiation rate:** weighted observed starts divided by weighted observed",
  "calendar woman-months, multiplied by 1,000. This describes how frequently",
  "method-use episodes begin during observed calendar time.\n"
)
```

The initiation analysis is implemented in `code/analyze_initiation.R`.

# Discontinuation estimation

## All-cause discontinuation

All discontinuation reasons are treated as a single endpoint. The cumulative
probability by month *t* is the weighted life-table quantity `1 - S(t)`.

## Discontinuation by reason

Reason-specific results use a weighted Aalen-Johansen competing-risk life
table. At each month, the weighted number of events for each reason is divided
by the weighted risk set. The cause-specific cumulative incidence is updated
using survival immediately before that month.

```{r competing-risk-example, eval=FALSE}
cause_increment <- survival_before_month *
  weighted_cause_events / weighted_risk_set

cumulative_incidence[cause] <-
  cumulative_incidence[cause] + cause_increment

survival_after_month <- survival_before_month *
  (1 - weighted_all_cause_events / weighted_risk_set)
```

The nine cause-specific cumulative incidences sum to the all-cause
discontinuation probability, subject only to floating-point precision.
Switching is calculated separately because switching can overlap with a stated
reason and therefore must not be added to the competing causes.

# Distribution of reasons among discontinuations

For each method, reason, and horizon:

```{r distribution-formula, echo=FALSE, results='asis'}
cat(
  "**Reason distribution (%) = reason-specific cumulative incidence /",
  "any-reason cumulative incidence x 100**"
)
```

This is a life-table-adjusted composition of discontinuation reasons, not a
simple unweighted cross-tabulation of observed events.

# Time to contraceptive reinitiation

The reinitiation sample begins with episodes discontinued for reasons other
than method failure/pregnancy while using or desire to become pregnant.

- Event of interest: first subsequent contraceptive method-use episode.
- Competing event: a subsequent pregnancy, birth, or termination before
  reinitiation.
- Censoring: interview reached before either event.
- Time scale: calendar months after the discontinued episode ends. A method
  beginning in the immediately following month is reinitiation at month 1.

Weighted Aalen-Johansen cumulative incidence is estimated for every month from
1 through 12. The dashboard shows the complete month-by-month curves and exact
values, plus a compact summary at 1, 3, 6, 9, and 12 months. A subsequent start
may be the same method or a different method; method switching is a separate
classification. The implementation is in `code/analyze_reinitiation.R`.

For each month, the numerator contribution is the weighted number of first
reinitiations in that month and the denominator is the weighted risk set still
observed and free of both reinitiation and pregnancy at the start of that
month. The Aalen-Johansen cumulative incidence sums the month-specific
reinitiation hazards after multiplying each by the probability of remaining
event-free through the preceding month. It is therefore not a crude proportion
with one fixed denominator.

The post-discontinuation destination analysis partitions reinitiation into the
first method subsequently initiated. Its conditional destination matrix divides
each destination-specific cumulative incidence by the cumulative incidence of
reinitiation with any method at the same horizon. Rows therefore sum to 100%
among estimated reinitiations. A second matrix retains pregnancy and no-event
outcomes and sums to 100% among all eligible discontinued episodes.

The gap-duration distribution uses the same weighted Aalen-Johansen estimates
but converts the cumulative reinitiation curve into mutually exclusive time
bands by differencing successive milestones. Those five bands, pregnancy before
reinitiation by 12 months, and no observed event by 12 months sum to 100%. The
method-effectiveness transition separately classifies the first subsequent
method as higher tier, a different method in the same tier, the same method,
or a lower tier. This is a descriptive programmatic classification, not a
judgment about an individual's preferred or clinically appropriate method.

## Reason-linked post-discontinuation outcomes

The reason-linked analysis retains all nine DHS reason groups and asks what
happens after an episode ends. The starting denominator for each row is all
observed discontinued episodes assigned to that reason. At each follow-up
month, the denominator used by the estimator is the survey-weighted risk set
still observed and free of both reinitiation and pregnancy. Numerator
contributions are the weighted first reinitiations or first pregnancy-related
events occurring in that month. Interview is right censoring.

Survey-weighted Aalen-Johansen cumulative incidence is reported at 1, 3, 6,
and 12 months for reinitiation, pregnancy before reinitiation, and no observed
event. These three probabilities sum to 100% within every reason and horizon.
Method failure and desire for pregnancy are retained to show every reason, but
their reproductive context differs from potentially service-responsive
discontinuation and must be interpreted separately.

Programmatically, this analysis distinguishes reasons commonly followed by
prompt switching from reasons followed by prolonged gaps or pregnancy before
reinitiation. It can inform counseling, side-effect management, referral,
commodity continuity, and follow-up, but it does not establish that a stated
reason caused the subsequent outcome.

## Previous method by reason

The previous-method-by-reason analysis repeats the same competing-risk model
within each method–reason subgroup. The starting denominator is all observed
discontinued episodes in that cell; monthly risk-set denominators and event
numerators are defined exactly as above. The dashboard allows the user to
select outcome and horizon and suppresses cells with fewer than 25 eligible
episodes; cells with 25–49 episodes are flagged.

This interaction is relevant because an apparently similar reason can imply a
different service response depending on the discontinued method. The results
are descriptive associations rather than causal effects. Emergency
contraception is interpreted separately because it is episodic rather than a
continuously used method.

# Dashboard section-by-section reading guide

The final dashboard places a visible explanation before the results in every
section and an expanded reading guide immediately below each table or figure.
The following documentation records the rows, columns, denominators,
numerators, estimator, total identities, and programmatic use.

## Section 01: Observed method initiation and dynamic method mix

- **Rows and columns:** The annual-use table has calendar years as rows and
  annual average monthly prevalence as columns. Initiation tables show weighted
  observed starts, their distribution by method, and starts per 1,000 observed
  calendar woman-months. The method-by-year table has discontinuation method
  categories as rows and calendar years as columns.
- **Numerator:** For annual use, survey-weighted calendar woman-months coded as
  any contraceptive use or modern-method use. For initiation, survey-weighted
  observed starts for a specified method.
- **Denominator:** All valid observed calendar woman-months in the year for
  annual use; all weighted observed starts for initiation mix; weighted
  observed calendar woman-months for initiation rate. Discontinuation uses the
  separate life-table risk set described in Section 02.
- **Totals:** Mutually exclusive method shares add to 100%. Rates and counts do
  not add to 100%. Annual any-method use equals modern plus traditional method
  use before rounding. Cumulative discontinuation milestones must not be added.
- **Method and rationale:** Survey-weighted descriptive ratios are appropriate
  for calendar prevalence and observed starts. Annual calendar prevalence is
  an average across observed months, not the DHS interview-date current-use
  estimate. Life-table estimates are needed for discontinuation because episode
  follow-up and censoring differ.
- **Programmatic relevance:** Distinguishes uptake from continuity. High
  initiation with rapid stopping may suggest a need for better method-fit
  counseling, resupply, or switching support rather than low demand.

## Section 02: All-cause discontinuation and monthly hazards

**Population used for the primary estimates:** These are overall all-cause
discontinuation estimates for the DHS-style eligible episode population. They
are not restricted to episodes initiated inside the 3-62-month analysis
window. Eligible episodes observed at risk during the window can contribute,
including delayed-entry episodes that began earlier within the recorded
calendar and continued into the window. Episodes already underway at the
oldest recorded calendar boundary are excluded because their true initiation
date and complete duration are unknown. Section 14 provides the separate,
narrower sensitivity analysis restricted to episodes whose starts and
immediately preceding calendar states are observed inside the analysis period.

- **Rows and columns:** Method rows show cumulative discontinuation at 3, 6, 9,
  and 12 months. Monthly columns show the conditional probability of stopping
  among episodes still continuing at that month's start.
- **Numerator:** Weighted all-cause discontinuations during month *t*.
- **Denominator:** Weighted eligible episodes continuing and observable at the
  start of month *t*, with delayed entry and censoring applied.
- **Totals:** Cumulative milestones and monthly hazards must not be added.
  Monthly risk sets change and survival compounds multiplicatively.
- **Method and rationale:** A weighted discrete-time life table estimates
  monthly hazards and `1 - S(t)` without imposing a fitted hazard shape.
- **Programmatic relevance:** Identifies methods and months where counseling,
  follow-up, side-effect support, or resupply may be most valuable.

## Section 03: Main reasons at each time point

- **Rows and columns:** Each method row lists the three largest reason-specific
  cumulative incidences at each milestone.
- **Numerator:** Weighted discontinuations for a reason, accumulated with the
  probability of having remained event-free beforehand.
- **Denominator:** Weighted episodes still at risk of discontinuation for any
  reason at each contributing month.
- **Totals:** The three displayed leaders do not necessarily total Any reason;
  other causes are omitted. Values across milestones must not be added.
- **Method and rationale:** Weighted Aalen-Johansen competing-risk cumulative
  incidence is used because an episode has one first recorded stopping reason.
- **Programmatic relevance:** Prioritizes the most prominent reasons by method
  and duration.

## Section 04: DHS-format discontinuation by reason and method

- **Rows and columns:** Rows are DHS aggregate method categories. Columns are
  nine mutually exclusive reasons, Any reason, switching, and episode support.
- **Numerator:** Weighted discontinuations for a reason during each month up to
  the selected horizon.
- **Denominator:** Weighted method episodes still at risk at the month's start.
- **Totals:** Nine reason CIFs add to Any reason. Switching must not be added
  because it can overlap with a stated reason.
- **Method and rationale:** Weighted Aalen-Johansen competing-risk estimation
  mirrors the DHS Table 7.11 structure while accommodating censoring.
- **Programmatic relevance:** Enables published-report validation and compares
  absolute reason-specific probabilities across methods.

## Section 05: Cumulative incidence and timing by reason

- **Rows and columns:** Method cards show reason-specific CIFs; monthly heat
  maps show cause-specific hazards; timing tables show the share of each
  12-month reason-specific CIF accumulated by months 3, 6, and 9.
- **Numerator:** Weighted events for the selected cause and month. For timing
  shares, the numerator is the cause's CIF at the earlier milestone.
- **Denominator:** The all-cause monthly risk set for hazards and that same
  cause's 12-month CIF for timing shares.
- **Totals:** Reason CIFs add to Any reason within a method and horizon. Monthly
  hazards and nested timing shares must not be added across time.
- **Method and rationale:** Aalen-Johansen CIFs preserve competing risks;
  nonparametric hazards preserve observed monthly timing.
- **Programmatic relevance:** Shows what reasons matter and when support must
  occur.

## Section 06: Distribution of reasons among discontinuations

- **Rows and columns:** The method view has reasons in rows and horizons in
  columns. The time-point view has methods in rows and reasons in columns.
- **Numerator:** Cause-specific CIF for one reason.
- **Denominator:** Any-reason CIF for the same method and horizon.
- **Totals:** Nine mutually exclusive reason shares total 100% within every
  method and horizon.
- **Method and rationale:** Ratios of weighted competing-risk CIFs retain
  consistency with the life-table results rather than raw event counts.
- **Programmatic relevance:** Shows the relative composition of discontinuation
  and therefore the balance of responses likely to be needed.

## Section 07: Time to contraceptive reinitiation

- **Rows and columns:** Method rows describe the method that ended. Reinitiation
  is shown cumulatively at 1, 3, 6, 9, and 12 months; pregnancy first and no
  event are shown at 12 months. Curves provide monthly values.
- **Starting population:** Discontinued episodes excluding method failure and
  desire for pregnancy.
- **Numerator:** Weighted first reinitiations or first pregnancy-related events
  in month *t*.
- **Denominator:** Weighted eligible episodes still observed and free of both
  outcomes at the beginning of month *t*.
- **Totals:** At one horizon, reinitiation + pregnancy first + no event = 100%.
  Reinitiation milestones are nested and must not be added.
- **Method and rationale:** Weighted Aalen-Johansen competing-risk CIF handles
  pregnancy as a competing event and interview censoring.
- **Programmatic relevance:** Quantifies protection gaps and opportunities for
  same-day switching and rapid follow-up.

## Section 08: Method used upon reinitiation

- **Rows and columns:** Rows are discontinued methods and columns are first
  subsequent methods. The main matrix is conditional among reinitiators; the
  expanded matrix uses all eligible episodes and adds pregnancy and no event.
- **Numerator:** Destination-specific Aalen-Johansen CIF.
- **Denominator:** Any-method reinitiation CIF for the conditional matrix; all
  eligible episodes for the expanded matrix.
- **Totals:** Destinations total 100% in the conditional matrix. Destinations +
  pregnancy + no event total 100% in the expanded matrix.
- **Method and rationale:** A multi-cause Aalen-Johansen partition preserves
  competing outcomes and censoring before conditional normalization.
- **Programmatic relevance:** Shows which replacement methods must be available
  and whether destination mix differs for early and later reinitiators.

## Section 09: Gap duration after discontinuation

- **Rows and columns:** Rows are discontinued methods. Columns are mutually
  exclusive restart intervals, pregnancy before restart, and no observed event.
- **Numerator:** Increment in reinitiation CIF within an interval or the
  12-month pregnancy/no-event state probability.
- **Denominator:** All eligible non-pregnancy-intent discontinued episodes in
  the method row, handled through monthly risk sets.
- **Totals:** Five restart intervals + pregnancy + no event total 100%.
- **Method and rationale:** Differences between Aalen-Johansen CIF milestones
  produce mutually exclusive gap categories while retaining competing risk and
  censoring.
- **Programmatic relevance:** Separates prompt switching from delayed restart
  and identifies groups with prolonged gaps.

## Section 09A: Integrated post-discontinuation protection pathways

- **Unit and denominator:** Eligible discontinued modern-method episodes; one
  woman may contribute multiple episodes. Monthly denominators are the
  survey-weighted episodes still observed and free of every competing pathway.
- **Mutually exclusive columns:** Immediate modern reinitiation in month 1;
  temporary non-use followed by modern reinitiation in months 2-3; a prolonged
  gap followed by modern reinitiation in months 4-12; first traditional-method
  use; pregnancy first; or no reinitiation or pregnancy observed by month 12.
  These six columns sum to 100% before rounding.
- **Method:** Survey-weighted multi-cause Aalen-Johansen cumulative incidence,
  with interview as right censoring. The first qualifying outcome determines
  the pathway.
- **Timing rationale:** Multi-country DHS research commonly defines prompt
  switching as switching within three months because most switching occurs in
  that period. Months 4-12 are retained as an operational prolonged-gap band,
  not a universal clinical threshold. Section 09 preserves the finer 4-6,
  7-9, and 10-12-month intervals.
- **Programmatic relevance:** Ordinary discontinuation combines immediate
  successful transitions with short gaps, prolonged gaps, movement to
  traditional methods, pregnancy, and no observed return. Separating these
  pathways identifies different needs for same-day switching, follow-up,
  counseling, method availability, and re-engagement.

## Section 10: Method-effectiveness transition

- **Rows and columns:** Rows are discontinued methods. Columns classify the
  subsequent method as higher tier, different same-tier method, same method, or
  lower tier, while retaining pregnancy and no event.
- **Numerator:** Destination-specific CIF grouped into the transition category.
- **Denominator:** All eligible discontinued episodes in the row method.
- **Totals:** All transition, pregnancy, and no-event columns total 100%.
- **Method and rationale:** Aalen-Johansen destination probabilities are
  aggregated with a prespecified effectiveness-tier mapping.
- **Programmatic relevance:** Summarizes pathways but does not determine whether
  a method is personally preferable or clinically suitable.

## Section 11: Outcomes after discontinuation by reason

- **Rows and columns:** Rows are all nine reasons. At 1, 3, 6, and 12 months,
  columns show reinitiation, pregnancy before reinitiation, and no event. Heat
  maps display the same values separately by outcome.
- **Numerator:** Weighted first reinitiations or pregnancy-related events in
  each month; no event is the remaining state probability.
- **Denominator:** Weighted episodes stopped for the reason and still free of
  both outcomes at each month's start.
- **Totals:** Three outcomes total 100% within each reason and horizon. Horizons
  are cumulative and must not be added.
- **Method and rationale:** Reason-stratified weighted Aalen-Johansen estimation
  accounts for competing pregnancy and censoring.
- **Programmatic relevance:** Connects why use ended with whether prompt
  switching, pregnancy, or prolonged non-use followed.

## Section 12: Previous method by reason

- **Rows and columns:** Rows are previous methods and columns are reasons. The
  controls select the cumulative outcome and horizon displayed.
- **Numerator:** Weighted first events for the selected outcome within the
  method-reason subgroup.
- **Denominator:** Weighted subgroup episodes still observed and free of both
  outcomes at each month's start.
- **Totals:** Within one subgroup and horizon, reinitiation + pregnancy first +
  no event = 100%. Cells across reasons must not be added.
- **Method and rationale:** Separate weighted Aalen-Johansen estimates are
  calculated within each subgroup. Fewer than 25 episodes are suppressed and
  25–49 are flagged.
- **Programmatic relevance:** Shows whether the same reason leads to different
  pathways depending on the method that ended.

## Section 13: Calendar use and initiation validation

- **Rows and columns:** Interview-year rows compare the v312 survey question
  with the vcal_1 state in the same interview month. A separate annual table
  reports calendar-based use by historical calendar year, followed by the
  method-by-year initiation mix.
- **Numerator:** Weighted women using contraception at interview for the
  current-use comparison; weighted use-coded woman-months for annual calendar
  prevalence; weighted observed starts for annual initiation mix.
- **Denominator:** Women with valid current-use and interview-month calendar
  fields for the paired comparison; valid observed woman-months within the year
  for annual prevalence; all observed weighted starts within the year for
  initiation mix.
- **Totals:** Any-method calendar use equals modern plus traditional use. Each
  year's five initiation-method shares add to 100% before rounding.
- **Method and rationale:** v312 is cross-sectional and measured only at
  interview, so it cannot generate historical annual estimates. The valid
  like-for-like check compares v312 with the calendar in that same interview
  month; retrospective annual trends remain calendar-based.
- **Programmatic relevance:** Confirms consistency of the two current-use
  measures while preserving the calendar's distinct value for examining change
  over historical years.

## Section 14: Observed-start cohort and contraceptive recovery

- **Primary DHS-style population:** Eligible episodes observed at risk inside
  the analysis window. Episodes beginning earlier within the recorded calendar
  can enter through delayed entry when they remain in use. Episodes already in
  progress at the first recorded calendar month are excluded because their
  start is not observed.
- **Observed-start cohort:** Episodes whose initiation and immediately
  preceding calendar state are observed. This narrower cohort links uptake to
  subsequent discontinuation and is the appropriate complementary leaky-bucket
  view.
- **Switching versus reinitiation:** Any-method reinitiation includes a return
  to the same method and a start of a different method. Switching is restricted
  to a first subsequent method different from the method that ended.
- **Post-discontinuation outcomes:** Different-method switching, same-method
  restarting, pregnancy before reinitiation, and no observed event form a
  competing-risk partition at each milestone and sum to 100% before rounding.
- **Statistical methods:** Survey-weighted discrete-time life tables estimate
  discontinuation. Survey-weighted Aalen-Johansen cumulative incidence
  estimates post-discontinuation pathways while treating pregnancy as a
  competing event and interview as right censoring.
- **Programmatic relevance:** The primary analysis remains comparable with the
  DHS report. The observed-start cohort adds a program-flow view of new method
  starts, loss through discontinuation, and recovery through same-method return
  or method switching.

## Section 15: How to interpret the episode cohorts and time horizons

- **Primary DHS-style discontinuation population:** Eligible contraceptive
  episodes observed at risk in the 3-62 months-before-interview window. This
  includes observed starts and eligible episodes that began before the
  62-month analysis boundary but remained in use inside the window and enter
  through delayed entry. It does not include episodes already underway at the
  oldest recorded calendar month because their initiation date and full
  duration are unknown.
- **Observed-start sensitivity population:** Episodes beginning 3-62 months
  before interview whose immediately preceding calendar state is also
  observed. The approximately five-year window selects eligible starts; it is
  not the duration of follow-up for the outcome.
- **Discontinuation clock:** Time zero is method initiation. Survey-weighted
  discrete-time life tables estimate cumulative discontinuation by months 3,
  6, 9, and 12 of use. Each month's numerator is weighted discontinuations in
  that month and its denominator is the weighted set still observed, using the
  method, and at risk at the beginning of the month. Incomplete observation is
  handled through right-censoring.
- **Post-discontinuation clock:** Time zero is the end of an eligible
  discontinued episode. Survey-weighted Aalen-Johansen cumulative incidence
  estimates different-method switching, same-method restarting, and pregnancy
  first through months 1, 3, 6, 9, and 12. The monthly denominator is the
  weighted set still observed and free of both reinitiation and pregnancy.
- **Outcome identity:** At each post-discontinuation horizon, different-method
  switching, same-method restarting, pregnancy first, and no observed event
  sum to 100% before rounding. Any-method reinitiation equals different-method
  switching plus same-method restarting.
- **Why both populations matter:** The primary construction preserves DHS
  comparability. The narrower observed-start cohort provides a complementary
  leaky-bucket view linking visible starts to first-year discontinuation and
  subsequent contraceptive recovery.

# Execute the reproducible pipeline

The chunk below runs the same modular scripts used to generate the final HTML.
It is intentionally hidden from the rendered documentation because the
preceding sections explain the transformations and estimators directly.

```{r run-analysis, include=FALSE}
analysis_log <- capture.output({
  source(file.path(code_dir, "replicate_table_7_11.R"), chdir = FALSE)
  source(file.path(code_dir, "analyze_monthly_hazards.R"), chdir = FALSE)
  source(file.path(code_dir, "analyze_initiation.R"), chdir = FALSE)
  source(file.path(code_dir, "analyze_reinitiation.R"), chdir = FALSE)
  source(file.path(code_dir, "analyze_leaky_bucket.R"), chdir = FALSE)
  source(
    file.path(code_dir, "analyze_reason_post_discontinuation.R"),
    chdir = FALSE
  )
  source(
    file.path(code_dir, "analyze_post_discontinuation_transitions.R"),
    chdir = FALSE
  )
  source(file.path(code_dir, "summarize_predominant_reasons.R"), chdir = FALSE)
  source(file.path(code_dir, "build_html_report.R"), chdir = FALSE)
  source(file.path(code_dir, "build_programmatic_summary.R"), chdir = FALSE)
})
```

# Selected validation outputs

```{r display-initiation}
initiation_mix <- read.csv(
  file.path(derived_dir, "observed_method_initiation_mix.csv"),
  check.names = FALSE
)
knitr::kable(
  initiation_mix[, c(
    "Method", "Observed starts (weighted)", "Share of observed starts (%)"
  )],
  digits = 1,
  caption = "Observed method initiation mix"
)

annual_use <- read.csv(
  file.path(derived_dir, "annual_calendar_contraceptive_use.csv"),
  check.names = FALSE
)
knitr::kable(
  annual_use[, c(
    "Calendar year", "Any contraceptive method use (%)",
    "Modern contraceptive method use (%)", "Coverage"
  )],
  digits = 1,
  caption = "Annual calendar-based contraceptive use"
)

annual_initiation <- read.csv(
  file.path(derived_dir, "annual_method_initiation_percent.csv"),
  check.names = FALSE
)
knitr::kable(
  annual_initiation,
  digits = 1,
  caption = "Annual method initiation mix"
)

current_use_comparison <- read.csv(
  file.path(derived_dir, "calendar_vs_v312_current_use.csv"),
  check.names = FALSE
)
knitr::kable(
  current_use_comparison,
  digits = 2,
  caption = "Current use at interview: calendar versus survey question"
)

leaky_bucket <- read.csv(
  file.path(derived_dir, "leaky_bucket_discontinuation_comparison.csv"),
  check.names = FALSE
)
knitr::kable(
  leaky_bucket,
  digits = 1,
  caption = "DHS-style versus observed-start discontinuation"
)

switching_pathways <- read.csv(
  file.path(derived_dir, "reinitiation_switching_pathways.csv"),
  check.names = FALSE
)
knitr::kable(
  subset(switching_pathways, Method == "All methods"),
  digits = 1,
  caption = "Timing of switching and reinitiation, all methods"
)

modern_protection_pathways <- read.csv(
  file.path(
    derived_dir,
    "post_discontinuation_modern_protection_pathways.csv"
  ),
  check.names = FALSE
)
knitr::kable(
  modern_protection_pathways,
  digits = 1,
  caption = "Integrated 12-month modern-protection pathways"
)
```

```{r display-discontinuation}
all_cause <- read.csv(
  file.path(derived_dir, "all_cause_life_table_published_groups.csv"),
  check.names = FALSE
)
knitr::kable(
  all_cause[, c("Method", "3 months", "6 months", "9 months", "12 months")],
  digits = 1,
  caption = "Cumulative all-cause discontinuation"
)
```

```{r validation-checks}
reason_results <- read.csv(
  file.path(derived_dir, "competing_risk_reasons_3_6_9_12_months.csv"),
  check.names = FALSE
)
reinitiation_results <- read.csv(
  file.path(derived_dir, "reinitiation_competing_risk_1_3_6_9_12.csv"),
  check.names = FALSE
)
gap_results <- read.csv(
  file.path(derived_dir, "post_discontinuation_gap_distribution.csv"),
  check.names = FALSE
)
effect_results <- subset(
  read.csv(
    file.path(derived_dir, "post_discontinuation_effectiveness_transition.csv"),
    check.names = FALSE
  ),
  Month == 12
)
destination_results <- read.csv(
  file.path(
    derived_dir,
    "post_discontinuation_destination_among_reinitiators.csv"
  ),
  check.names = FALSE
)
reason_outcome_results <- read.csv(
  file.path(derived_dir, "reason_post_discontinuation_outcomes.csv"),
  check.names = FALSE
)
method_reason_outcome_results <- read.csv(
  file.path(
    derived_dir,
    "reason_previous_method_post_discontinuation_outcomes.csv"
  ),
  check.names = FALSE
)
modern_protection_results <- read.csv(
  file.path(
    derived_dir,
    "post_discontinuation_modern_protection_pathways.csv"
  ),
  check.names = FALSE
)

reason_columns <- c(
  "Method failure", "Desire to become pregnant",
  "Other fertility-related reasons", "Changes in menstrual bleeding",
  "Other side effects/health concerns", "Wanted more effective method",
  "Other method-related reasons", "Husband/partner disapproved",
  "Other reasons"
)

reason_identity_error <- max(abs(
  rowSums(reason_results[, reason_columns]) - reason_results$`Any reason`
))
reinitiation_identity_error <- max(abs(
  reinitiation_results$`Reinitiated (%)` +
    reinitiation_results$`Pregnancy competing event (%)` +
    reinitiation_results$`No observed event (%)` - 100
))
gap_identity_error <- max(abs(rowSums(gap_results[, -1]) - 100))
effect_identity_error <- max(abs(
  rowSums(effect_results[, setdiff(names(effect_results), c(
    "Discontinued method", "Month"
  ))]) - 100
))
destination_identity_error <- max(abs(
  rowSums(destination_results[, setdiff(
    names(destination_results),
    c("Discontinued method", "Month")
  )]) - 100
))
reason_outcome_columns <- c(
  "Reinitiated (%)", "Pregnancy before reinitiation (%)",
  "No observed event (%)"
)
reason_outcome_identity_error <- max(abs(
  rowSums(reason_outcome_results[, reason_outcome_columns]) - 100
))
method_reason_identity_error <- max(abs(
  rowSums(method_reason_outcome_results[, reason_outcome_columns]) - 100
))
switching_pathway_identity_error <- max(abs(
  switching_pathways$`Switched to different method (%)` +
    switching_pathways$`Restarted same method (%)` +
    switching_pathways$`Pregnancy before reinitiation (%)` +
    switching_pathways$`No observed event (%)` - 100
))
switching_restart_identity_error <- max(abs(
  switching_pathways$`Switched to different method (%)` +
    switching_pathways$`Restarted same method (%)` -
    switching_pathways$`Any-method reinitiation (%)`
))
annual_use_identity_error <- max(abs(
  annual_use$`Any contraceptive method use (%)` -
    annual_use$`Modern contraceptive method use (%)` -
    annual_use$`Traditional method use (%)`
))
annual_initiation_identity_error <- max(abs(
  colSums(annual_initiation[, -1, drop = FALSE]) - 100
))
modern_protection_columns <- c(
  "Continued modern protection",
  "Temporary non-use then modern reinitiation",
  "Prolonged gap then modern reinitiation",
  "Traditional-method use",
  "Pregnancy first",
  "No reinitiation or pregnancy observed by 12 months"
)
modern_protection_identity_error <- max(abs(
  rowSums(modern_protection_results[, modern_protection_columns]) - 100
))

data.frame(
  Check = c(
    "Sum of reason-specific CIFs equals any-reason discontinuation",
    "Reinitiation, pregnancy, and event-free components sum to 100%",
    "Gap-duration outcome rows sum to 100%",
    "Effectiveness-transition outcome rows sum to 100%",
    "Conditional destination rows sum to 100%",
    "Reason-linked outcome rows sum to 100%",
    "Method-by-reason outcome rows sum to 100%",
    "Annual any-method use equals modern plus traditional use",
    "Annual initiation method shares sum to 100%",
    "Switching pathway components sum to 100%",
    "Different-method plus same-method restart equals any reinitiation",
    "Integrated modern-protection pathway rows sum to 100%"
  ),
  `Maximum absolute error` = c(
    reason_identity_error,
    reinitiation_identity_error,
    gap_identity_error,
    effect_identity_error,
    destination_identity_error,
    reason_outcome_identity_error,
    method_reason_identity_error,
    annual_use_identity_error,
    annual_initiation_identity_error,
    switching_pathway_identity_error,
    switching_restart_identity_error,
    modern_protection_identity_error
  ),
  Status = c(
    ifelse(reason_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(reinitiation_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(gap_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(effect_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(destination_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(reason_outcome_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(method_reason_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(annual_use_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(annual_initiation_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(switching_pathway_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(switching_restart_identity_error < 1e-8, "PASS", "FAIL"),
    ifelse(modern_protection_identity_error < 1e-8, "PASS", "FAIL")
  ),
  check.names = FALSE
) |>
  knitr::kable(digits = 12, caption = "Internal validation checks")
```

# Rebuilding the final HTML

Rendering this R Markdown file executes the complete analysis and recreates:

```text
output/ethiopia_discontinuation_3_6_9_12.html
```

From RStudio, use **Knit**, or run:

```{r render-command, eval=FALSE}
rmarkdown::render("ethiopia_discontinuation_analysis.Rmd")
```

The modular source files remain in `code/` so that individual analysis stages
can be reviewed and tested independently.
